\chapter{Limitations}
\label{sec:limitations}
Four limitations bound our results.

\section{Pitch and chin-down control}
\label{sec:lim-pitch}
The one axis our pipeline cannot fix is chin-down rotation; it is also the axis
\emph{every} backbone in Table~\ref{tab:backbone} fails, and chin-rotated
controls recover chin-up but not chin-down. We treat it as a learned prior of
the distilled backbone rather than a deficiency of the method.

\section{Scale and backbone constraints}
\label{sec:lim-scale}
Our learned allocator works because it acts
inside the existing attention budget; attempts to train a fresh, higher-capacity
identity pathway did not generalize at our identity count, leaving open whether
that is a data-scale or a distillation effect (the few-step backbone runs
off-distribution under heavy conditioning).

\section{The one-shot ceiling}
\label{sec:lim-oneshot}
A single frontal reference fundamentally under-determines the far side of a profile
face; a small fraction of identities are simply hard for the backbone to render
regardless of conditioning, bounding achievable fidelity.

\section{The pose--identity trade-off is fundamental, not fully solved}
\label{sec:lim-tradeoff}
Instruction \emph{image-editing} models are an appealing alternative---they edit
the reference face directly rather than re-synthesizing it---yet they inherit the
\emph{same} tension we study: Step1X-Edit~\cite{step1xedit} preserves identity but
barely rotates ($11.4\%$ yaw), whereas OmniGen2~\cite{omnigen2} rotates
substantially ($44.4\%$ yaw) but at a markedly lower identity ($0.308$ AdaFace),
trading one objective for the other rather than achieving both. This indicates the
trade-off is not an artifact of identity \emph{adapters} but a general property of
driving a frozen generator toward off-distribution pose. Our $\gamma$-gate is, to
our knowledge, the only mechanism evaluated here that improves identity \emph{at no
pose cost}, but it does not eliminate the trade-off---identity at profile still
falls below frontal reconstruction (Sec.~\ref{sec:lim-oneshot})---and a method that
combines direct pixel editing with attention-budget reallocation may push the
frontier further than either alone.

