Recent identity-preserving personalization methods can faithfully reproduce a reference face~\cite{ipadapter,photomaker,instantid,pulid,uniportrait}, but they remain strongly biased toward near-frontal generations. Under \emph{posture-requested} instructions---turning to the side, looking over the shoulder, or tilting the chin up or down---existing methods often collapse back to a frontal view, effectively ignoring the requested posture. We study \emph{posture-requested personalized text-to-image generation}: given a \emph{single} reference identity and a textual posture request, generate that identity at the requested posture---turned left or right, optionally over the shoulder, chin up or down---and, just as critically, \emph{measure} the result with metrics that are not themselves biased toward frontal faces.

We present \textbf{Rotate-ID}, a \emph{one-shot} system built on a few-step rectified-flow DiT (FLUX.2-klein). A structural pose control, supplied \emph{in context}, drives the requested head turn; identity is preserved by re-allocating the frozen backbone's attention toward the reference---a mechanism we verify causally (identity and pose compete for one shared attention budget) and distill into a $\sim$200-parameter learned gate (the $\gamma$-gate) that refines identity \emph{at no pose cost}, with the prompt unchanged.

The second half of the problem is measurement: face encoders read identity reliably near frontal but \emph{under}-report it as the \emph{same} person turns away from frontal---so merely staying frontal scores high identity---and CLIP alignment cannot tell whether the posture was realized. Our \emph{pose-rectified performance metrics} close both gaps: an explainable 3D pose check verifies each requested turn, a pose-adapted read-out scores identity with yaw-matched recognizers, and a \emph{pose-aware} score credits identity only at the posture actually produced. The suite is instantiated as a benchmark of $40$ posture requests over eight pose categories.

Under these pose-rectified metrics, Rotate-ID substantially improves pose controllability over PhotoMaker-V2, the strongest identity-personalization baseline, more than doubling its yaw accuracy ($82.6\%$ vs.\ $36.1\%$). It further realizes over-the-shoulder rotations at measured yaw angles no baseline attains, maintaining usable identity in the large-rotation regime where competing methods rarely produce valid rotated samples; under the pose-aware identity, Rotate-ID is the strongest tuning-free method, while the frontal-collapsing baselines' apparent identity lead disappears once pose is taken into account.
